---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Predicting Salary of Major League Baseball Players Using Nonparametric Methods

This repository consists of an R project for MAST90111 - Advanced Statistical Modelling, holding the source files for reproducing the deliverable, including the code files and their version history.

## Requirements

To reproduce the deliverables of the project, you will need to install the following packages and their dependencies as follows:

```{r lib, eval = FALSE}
install.packages("tidyverse")
install.packages("xaringan")
install.packages("xaringanthemer")
tinytex::install_tinytex()
```

## Reproducible research

The main deliverable of the project consists of a PDF report rendered by R Markdown.

```{r render-pdf, eval = FALSE}
rmarkdown::render("report/report.Rmd")
```

The presentation of the project consists of an HTML slide rendered by R Markdown with the [xaringan](https://github.com/yihui/xaringan) package.

```{r render-slide, eval = FALSE}
rmarkdown::render("slide/slide.Rmd")
```

The [report](https://github.com/szmsu2011/mast90111proj/tree/main/report) directory consists of source files for reproducing the PDF report. The [slide](https://github.com/szmsu2011/mast90111proj/tree/main/slide) directory consists of source files for reproducing the HTML slide.
